\documentclass{article}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage[square, numbers, sort&compress]{natbib}

\bibliographystyle{abbrvnat}

\title{\texttt{bigrig}: A historical biogeography range simulator using the
	DEC[+J] model}
\author{Ben Bettisworth}

\begin{document}

\newcommand{\CountFull}[1]{|#1|_\text{full}}
\newcommand{\CountEmpty}[1]{|#1|_\text{empty}}

\maketitle

\section{Terminology}

We denote the number of regions for a range \( r \) as \( |r| = n\).
Each region can either be occupied, in which case we say it is full, or it can
be unoccupied, in which case we say it is empty.
We write the number of full regions a region has as \( \CountFull{r} \), and
the number of empty regions as \( \CountEmpty{r} \).

\section{Methods}

\subsection{The DEC[+J] Model} \label{sec:model}

The Dispersion, Extinction and Cladogenesis (DEC) model defines the probability
of observing particular geographical history given a pair of parameters for
dispersion and extinction rates.
In the model, there are 3 processes affecting the range of a particular
species.
Dispersion and extinction are the processes by which a range is added or
removed over time due to random drifts.
The process is modeled as each region independently transitioning to a
different state (either from full to empty or vise versa) with waiting times
distributed exponentially based on the respective rate.
This is to say, for a region \( r \) which is full in a model with extinction
rate \( e \) the waiting time \( w \) for \( r \) to transition to empty is \(
w \sim \text{Exp}(e) \).

The third process is cladogenesis, I.E. when a parent species splits into two
daughter species.
When this occurs, some regions from the parent's range will be inherited by one
daughter, and rest will be inherited by the other daughter.
The particular way in which the parent ranges are inherited is governed by a
set of scenarios.
In the original Ree \textit{et. al.}\cite{ALikelihoodFrReeR2005} paper, these
scenarios are simply labeled Scenario 1, 2, and 3.
However, the scenarios are intended to represent realistic cladogenesis
methods.
In the interest of clarity and memorability, we will use the more common names,
similar to how Matkze \cite{ModelSelectionMatzke2014} has done in his paper.

The three cladogenesis scenarios possible in the original DEC model are:
allopatry (or vicariance), where the daughter ranges are disjoint; sympatry,
where the daughter species share some ranges; and copy, where both daughter
ranges are the same as the parent range.
An assumption by the DEC model is that speciation can only occur in one range,
and therefore, at least one of the daughter ranges will consist of only a
single region.
Additionally, the copy scenario can only occur when the parent range is a
single region.
When a cladogenesis event occurs (that is, when a node splits into two
daughters) the DEC model computes the probability of these events as being
equivalent, where possible.
This is to say, when considering if a particular cladogenesis pattern was the
result of allopatry or sympatry, DEC considers these events to have equal
weight.
To compute the probability of a particular type of cladogenesis event over the
set \(T = \left\{\text{Allopatry, Sympatry, Copy}\right\}\) we compute
\[
	P(t |
	s) = \frac{C(t | s)}{\sum_{u \in T} C(u | s)},
\]
where \( C(t|s) \) is the
count of cladogenesis events of type $ t $ which are possible given the parent
range \( s \).

Matzke\cite{ModelSelectionMatzke2014} extended the cladogenesis probability
computation by: allowing for a new "jump" type of cladogenesis events; and
weights for each of the different cladogenesis types such that a particular
type can be more likely than the others.
This pair of modifications to the DEC model are generally denoted as DEC+J,
similar to how gamma rate categories is denoted GTR+G4 in phylogenetic models.
Under this modified model, the probability of a particular type of cladogenesis
event is computed
\[
	P(t | s) = \frac{w_t C(t | s)}{\sum_{u \in T} w_u C(u |
		s)},
\]
where $w_t$ is the weight for the cladogenesis type $t$.
Additionally, $T$ is augmented by the new type "jump" to become \(T =
\left\{\text{Allopatry, Sympatry, Copy, Jump}\right\} \).

\subsection{Design}

For the purposes of simulating a biogeographical history under the DEC[+J]
model I have broken up the model into two phases: the evolution of a range
\textit{along} a branch, which encompasses both extinction and dispersion; and
the splitting of a range at cladogenesis.
I call these two phases of simulation the spread and the split, respectively.

For both phases, I have implemented two independent methods of simulating
results, with the purpose of ensuring the accuracy of the simulation.
The first method of simulating results for both phases is a rejection method.
Rejection methods are slow, but reliable, methods of simulating stochastic
functions.
Of particular interest here is that they are easy to implement correctly, as
they are well understood, even for complex problems.
In general, they operate by generating a random result uniformly, and then
checking if that solution is valid.
If the solution is valid, the process returns the result.
Otherwise, it generates a new result and checks again, repeating until a valid
result is generated.

The other general method of simulating results that I have implemented is what
I call analytic methods.
These methods implement optimizations which ought to be true analytically, but
are more complex.
Additionally, an effort is made to only generate a single result, greatly
accelerating the simulation.
Often, methods like this are an order of magnitude or more faster than
rejection methods, depending on the problem.
However, this additional complexity can make it difficult to ensure accurate
results.
Therefore, by checking the results of the analytic methods against the results
of rejection methods, we get the best of both: an correctly implemented and
fast method of simulation.

\subsection{Simulating the Spread}

As mentioned in Section~\ref{sec:model}, the evolution of a range along a
branch is modeled as a Poisson process, analogous to the character evolution
models used in phylogenetic inference.

The problem is this: given an initial range \( r \) with \( |r| = n \) regions,
and a branch length \( t \), produce a final range which has undergone
extinction and dispersion at rates \( e \) and \( d \), respectively.
When a range experiences dispersion, it gains a region.
Analogously, when a range experiences extinction, it loses a region, Each of
the regions is treated as an independent process from each other.
This observation is the basis for the rejection method of spread simulation:
randomly generate \( n \) waiting times, labeled \( w_i \), for each region.
Specifically, the waiting times are distributed as
\begin{equation}
	\label{eq:exp-rejection} w_i \sim
	\begin{cases}
		\text{Exp}(e) & \text{if } r_i
		= 1                            \\ \text{Exp}(d) & \text{if } r_i = 0.
	\end{cases}
\end{equation}
Once all $n$ waiting times are generated, we pick the smallest, say $w_i$, and
invert $r_i$.
We repeat this process until the total waiting time exceeds the branch length,
at which point we return a list of transition events, excluding the final event
who's waiting time exceeded the branch length.

Keen readers might already be aware of the fact that if we have a set of
independent processes like above, the entire set can be represented by a single
exponential distribution, with \(w \sim \text{Exp}(t) \) where
\begin{equation}
	\label{eq:exp-param} t = e \times \CountFull{r} + d \times \CountEmpty{r}.
\end{equation}
This second method then is to compute \( t \) using Eq.~\ref{eq:exp-param},
draw a waiting waiting time \( w \sim \text{Exp}(t) \).
Once a time is rolled, we pick a region with weights equal to the exponential
distribution parameter in Eq.~\ref{eq:exp-rejection}.

We implement both of the above methods, but only use the second to preform
simulations.
The first method is used to check the results of the second method.

\subsection{Simulating the Split}

\bibliography{references}

\end{document}
